---
title: "March Madness Analysis"
author: "Tobin Haefele"
date: "sys.Date()"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# import libraries
library(tidyverse)
library(lubridate)
library(readxl)
library(ggplot2)
library(ggcorrplot)
library(rpart)
library(rpart.plot)

# import data
df <- read.csv("archive/cbb.csv")
```
## Introduction
In my analysis I want to examine the relationship between various statistics and the success of college basketball teams in the NCAA tournament. I will use the data to create a logistic regression model to predict the success of teams in the tournament. I will also use a classification tree and random forest to predict the success of teams in the tournament. I will then compare the models to see which is the most accurate.

### Data
I consolidated the data into a single csv file which allows for easy access and manipulation. The data includes over 10 years of college basketball statistics and tournament results. Mainly I want to focus on the teams that won the tournament and the statistics that are most important in determining the success of a team in the tournament.

================================================================================
## Initial regression analysis
My initial model will add a column for tournament champion as a binary. I will then create a logistic regression model to predict the success of teams in the tournament.
```{r}
###Regression analysis of tournament teams

#add a column for champion as a binary
df$CHAMPION <- ifelse(df$POSTSEASON == "Champions", 1, 0)

#remove columns that are not needed, all teams play the same number of games
df <- df %>% select(-c(G, POSTSEASON))

print(colnames(df))

#logistic regression model
logit_model <- glm(CHAMPION ~ ADJOE + ADJDE + BARTHAG + EFG_O + EFG_D,TOR + TORD + ORB + DRB + FTR + FTRD + `X2P_O` + `X2P_D` + `X3P_O` + `X3P_D` + ADJ_T + WAB, data = df, family = "binomial")

#summary of logistic regression model
summary(logit_model)
```
================================================================================
```{r}
#test a prediction model by creating a training and test set
#set seed
set.seed(123)

#split data into training and test set
train <- sample(1:nrow(df), 0.8*nrow(df))
train_set <- df[train,]
test_set <- df[-train,]

# create logistic regression model
logit_model2 <- glm(CHAMPION ~ ADJOE + ADJDE + EFG_O + EFG_D,TOR + TORD + ORB + DRB + FTR + FTRD + `X2P_O` + `X2P_D` + `X3P_O` + `X3P_D` + ADJ_T + WAB,data = train_set, family = "binomial")

# summary of logistic regression model
summary(logit_model)

#predict on test set
pred <- predict(logit_model2, test_set, type = "response")



```
================================================================================
```{r}
#create a classification tree
library(rpart)

#set seed
set.seed(123)

#split data into training and test set
train <- sample(1:nrow(df), 0.8*nrow(df))

train_set <- df[train,]
test_set <- df[-train,]

# create classification tree
tree <- rpart(CHAMPION ~ ADJOE + ADJDE + EFG_O + EFG_D,TOR + TORD + ORB + DRB + FTR + FTRD + `X2P_O` + `X2P_D` + `X3P_O` + `X3P_D` + ADJ_T + WAB, data = train_set, method = "class")

#plot tree with rpart.plot
library(rpart.plot)
rpart.plot(tree)

#summary of tree
summary(tree)

#predict on test set
pred <- predict(tree, test_set, type = "class")

#confusion matrix
table(test_set$MDE, pred)

#label confusion matrix
library(caret)
confusionMatrix(table(test_set$MDE, pred))

#accuracy
mean(test_set$MDE == pred)

#precision
mean(test_set$MDE == pred & pred == 1)

#recall
mean(test_set$MDE == pred & test_set$MDE == 1)

#f1 score
f1_score <- function(x, y){
  precision <- mean(x == y & y == 1)
  recall <- mean(x == y & x == 1)
  return(2 * precision * recall / (precision + recall))
}

f1_score(test_set$MDE, pred)

#ROC curve
library(ROCR)

pred <- prediction(pred, test_set$MDE)
perf <- performance(pred, "tpr", "fpr")
plot(perf)

#AUC
auc <- performance(pred, "auc")

#plot ROC curve with AUC
plot(perf)
abline(a = 0, b = 1)
text(0.5, 0.5, paste("AUC = ", round(auc, 2)))
```
================================================================================
```{r}
#random forest
library(randomForest)

#set seed
set.seed(123)

#split data into training and test set
train <- sample(1:nrow(df), 0.8*nrow(df))

train_set <- df[train,]
test_set <- df[-train,]

# create random forest
rf <- randomForest(MDE ~ ADJOE + ADJDE + DRB + ADJ_T, data = train_set, ntree = 100)

#summary of random forest
summary(rf)

#predict on test set
pred <- predict(rf, test_set, type = "class")

#confusion matrix
table(test_set$MDE, pred)

#label confusion matrix
library(caret)
confusionMatrix(table(test_set$MDE, pred))

#accuracy
mean(test_set$MDE == pred)

#precision
mean(test_set$MDE == pred & pred == 1)

#recall
mean(test_set$MDE == pred & test_set$MDE == 1)

#f1 score
f1_score <- function(x, y){
  precision <- mean(x == y & y == 1)
  recall <- mean(x == y & x == 1)
  return(2 * precision * recall / (precision + recall))
}

f1_score(test_set$MDE, pred)
```

### Glossary
Variables
RK (Only in cbb20): The ranking of the team at the end of the regular season according to barttorvik

TEAM: The Division I college basketball school

CONF: The Athletic Conference in which the school participates in (A10 = Atlantic 10, ACC = Atlantic Coast Conference, AE = America East, Amer = American, ASun = ASUN, B10 = Big Ten, B12 = Big 12, BE = Big East, BSky = Big Sky, BSth = Big South, BW = Big West, CAA = Colonial Athletic Association, CUSA = Conference USA, Horz = Horizon League, Ivy = Ivy League, MAAC = Metro Atlantic Athletic Conference, MAC = Mid-American Conference, MEAC = Mid-Eastern Athletic Conference, MVC = Missouri Valley Conference, MWC = Mountain West, NEC = Northeast Conference, OVC = Ohio Valley Conference, P12 = Pac-12, Pat = Patriot League, SB = Sun Belt, SC = Southern Conference, SEC = South Eastern Conference, Slnd = Southland Conference, Sum = Summit League, SWAC = Southwestern Athletic Conference, WAC = Western Athletic Conference, WCC = West Coast Conference)

G: Number of games played

W: Number of games won

ADJOE: Adjusted Offensive Efficiency (An estimate of the offensive efficiency (points scored per 100 possessions) a team would have against the average Division I defense)

ADJDE: Adjusted Defensive Efficiency (An estimate of the defensive efficiency (points allowed per 100 possessions) a team would have against the average Division I offense)

BARTHAG: Power Rating (Chance of beating an average Division I team)

EFG_O: Effective Field Goal Percentage Shot

EFG_D: Effective Field Goal Percentage Allowed

TOR: Turnover Percentage Allowed (Turnover Rate)

TORD: Turnover Percentage Committed (Steal Rate)

ORB: Offensive Rebound Rate

DRB: Offensive Rebound Rate Allowed

FTR : Free Throw Rate (How often the given team shoots Free Throws)

FTRD: Free Throw Rate Allowed

2P_O: Two-Point Shooting Percentage

2P_D: Two-Point Shooting Percentage Allowed

3P_O: Three-Point Shooting Percentage

3P_D: Three-Point Shooting Percentage Allowed

ADJ_T: Adjusted Tempo (An estimate of the tempo (possessions per 40 minutes) a team would have against the team that wants to play at an average Division I tempo)

WAB: Wins Above Bubble (The bubble refers to the cut off between making the NCAA March Madness Tournament and not making it)

POSTSEASON: Round where the given team was eliminated or where their season ended (R68 = First Four, R64 = Round of 64, R32 = Round of 32, S16 = Sweet Sixteen, E8 = Elite Eight, F4 = Final Four, 2ND = Runner-up, Champion = Winner of the NCAA March Madness Tournament for that given year)

SEED: Seed in the NCAA March Madness Tournament

YEAR: Season